Modeling HEPES buffer with a charge-balance approach

### 2023-11-17 manuscript.Rmd
Vectorize: differentiating past root-finding
It's actually possible, but takes certain gymnastics in both R and
calculus, to obtain the symbolic Jacobian despite a root-finding
operation in the middle.

>>>>

Unfortunately, the use of automatic vectorization precludes the use of
fully automatic symbolic differentiation: static code analysis will not
go past `vapply`, `mapply`, or `uniroot`. Still, it is possible to
construct most of the expression piece by piece and have it
differentiated this way:

```{r}
charge_balance <- local({
	XX <- quote(H^3/(10^-pK3*10^-pK2*10^-pK1)+H^2/(10^-pK3*10^-pK2)+H/(10^-pK3))
	IV <- substitute(HEPTOT/(1+XX))
	I <- substitute(IV*H^3/(10^-pK3*10^-pK2*10^-pK1))
	II <- substitute(IV*H^2/(10^-pK3*10^-pK2))
	III <- substitute(IV*H/10^-pK3)
	nlsr::codeDeriv(
		substitute(H - kw/H + SID + I*2 + II - IV),
		c('H', 'pK1', 'pK2', 'pK3')
	)
})
# a few screenfuls of code; >2500 pages of output with verbose=TRUE
```

In order to obtain the final Jacobian, we will have to somehow
differentiate past the root-finding operation. For that we will use the
implicit function theorem. Let the charge-balance equation be
$f([\mathrm H], \theta) = 0$, with $[\mathrm H]$ being the desired
equilibrium concentration of protons and $\theta$ symbolising all of the
parameters of the equation. It implicitly defines the function
$[\mathrm H]_\mathrm{eq}(\theta)$ that satisfies
$F(\theta) = f([\mathrm H]_\mathrm{eq}(\theta), \theta) = 0$. Differentiating the
latter equation by $\theta$, we get $\frac{dF}{d\theta} =
\frac{\partial{}f}{\partial[\mathrm H]}
\frac{d[\mathrm H]_\mathrm{eq}}{d\theta} +
\frac{\partial{}f}{\partial\theta} \frac{d\theta}{d\theta} = 0$, from
which we get $\frac{d[\mathrm H]_\mathrm{eq}}{d\theta} =
-\frac{\partial{}f/\partial\theta}{\partial{}f/\partial[\mathrm H]}$:

```{r}
pH.sym <- function(SID, HEPTOT, pK1, pK2, pK3) {
 H <- vapply(SID, function(SID) { # for every point...
  # find the root of the function
  H <- uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
                 HEPTOT = HEPTOT, SID = SID,
                 pK1 = pK1, pK2 = pK2, pK3 = pK3)$root
  # evaluate the derivatives by H and parameters
  g <- attr(eval(charge_balance), 'gradient')
  # and calculate the derivative of the H_eq by parameters;
  # combine together for easier access by vapply
  c(H, -g[,2:4]/g[,1])
 }, numeric(4)
 )
 # now take the negative logarithm to obtain pH
 structure(-log10(H[1,]), gradient = -(1/(H[1,] * log(10))) * t(H[-1,]))
}
pH.sym(
 c(.04, .035), .05, -1, 3, 7.55
)
```

This gets similar, but not exactly the same results as numeric
differentiation:

```{r}
nlsr::numericDerivR(
 quote(pHm(
  c(.04, .035), .05, pK1, pK2, pK3
 )),
 c('pK1', 'pK2', 'pK3'),
 list2env(list(pK1 = -1, pK2 = 3, pK3 = 7.55)),
 central = TRUE
)
```
